MaBoSS for HPC environments: Implementations of the continuous time Boolean model simulator for large CPU clusters and GPU accelerators

### 2024-02-14 Article.tex
giga format

\RequirePackage[2020-02-02]{latexrelease}
\documentclass[times, twoside]{zHenriquesLab-StyleBioRxiv}
%\usepackage{blindtext}
\usepackage{algorithm}
\usepackage{algcompatible}

% Please give the surname of the lead author for the running footer
\leadauthor{Šmelko}

@@

\title{MaBoSS for HPC environments: Implementations of the continuous time Boolean model simulator for large CPU clusters and GPU accelerators}
\shorttitle{MaBoSS for HPC environments}

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[1,\Letter]{Adam Šmelko}
\author[2]{Miroslav Kratochvíl}
\author[3,4,5]{Emmanuel Barillot}
\author[3,4,5]{Laurence Calzone}
\author[3,4,5,\Letter]{Vincent Noël}

\affil[1]{Department of Distributed and Dependable Systems, Charles University, Prague, Czech Republic}
\affil[2]{Luxembourg Centre for Systems Biomedicine, University of Luxembourg, Esch-sur-Alzette, Luxembourg}
\affil[3]{Institut Curie, Université PSL, F-75005, Paris, France}
\affil[4]{INSERM, U900, F-75005, Paris, France}
\affil[5]{Mines ParisTech, Université PSL, F-75005, Paris, France}
\maketitle

@@

\begin{keywords}
Computational Biology | High Performance Computing | Boolean models
\end{keywords}

\begin{corrauthor}
smelko@d3s.mff.cuni.cz, vincent.noel@curie.fr
\end{corrauthor}
\section*{Introduction}
Biological systems are large and complex, and understanding their behavior remains critical to designing new therapies for complex diseases such as cancer. A crucial approach in this endeavor is building computational models from existing knowledge and analyzing them to find intervention points and predict the efficacy of new treatments.
Many different frameworks have been used to describe biological systems, from quantitative systems of differential equations to more qualitative approaches such as Boolean models. While the former seems more adapted to represent complex behavior, such as non-linear dependencies, the latter is being increasingly used because it is capable of representing very large systems.
Many Boolean models have been built to describe biological systems to tackle a variety of problems: from understanding fundamental properties of cell cycle~\cite{faure2006cellcycle,sizek2019boolean} to more advanced properties of cancer~\cite{fumia_carcinogenesis_2013, montagud2022prostate}.

Historically, the task of building Boolean models involved reading an extensive amount of literature and summarizing it in a list of essential components and their interactions. More recently, thanks to advances in databases listing such interactions~\cite{licata2020signor,turei2016omnipath} and to experimental techniques providing information on a bigger number of components, the automatic methods have been designed to infer Boolean formulas from the constraints encoded in the knowledge and the experimental data~\cite{10.1093/bioinformatics/btaa484,chevalier2020synthesis,benevs2023boolean} to build large Boolean models. While this effort faces many challenges, we believe it is the best way to study the large-scale complexity of biological systems. However, in order to analyze the dynamic properties of such large Boolean models, we need to efficient simulation tools. In this effort, adapting existing software to modern HPC computing architectures is an important goal.

Our work focuses on MaBoSS~\cite{stoll2012continuous, stoll2017maboss}, a stochastic Boolean simulator that performs estimations of state probability trajectories based on Markov chains. 
This paper discusses porting the MaBoSS simulation algorithm to two hardware architectures: GPU accelerators and large CPU clusters with an MPI parallel implementation. The source codes are available at \url{https://github.com/sysbio-curie/MaBoSSG/} and \url{https://github.com/sysbio-curie/MaBoSS-env-2.0/}, together with associated benchmarking data.

\section*{Background}

@@

\section*{Complexity evaluation and parallelization opportunities of MaBoSS}

@@

\subsection*{Simulation complexity and parallelization}

@@

\subsection*{Statistics computation complexity and parallelization}

@@

\section*{Implementation}\label{sec:implementation}

@@

\subsection*{MaBoSS.GPU}
Porting MaBoSS algorithm to GPUs poses an interesting task from the optimizations perspective. A simple rewrite to an annotation-based tool, such as OpenACC, may not be the most optimal strategy due to the usage of complex data structures, such as hash maps and expression trees, that are not well suited for GPU hardware. We divided the implementation into two parts: the bare \emph{simulation} and the trajectory \emph{statistics computation}.

\subsubsection*{Simulation}

@@

\subsubsection*{Statistics computation}

@@

\subsection*{MaBoSS.MPI}
Since MaBoSS.MPI and MaBoSS.GPU projects were developed in parallel, the MPI implementation of MaBoSS is based on the original CPU code. Same as in GPU code, CPU implementation also assigns a single thread to a single trajectory. The MPI implementation simply extends this distribution of work among multiple MPI nodes, adding the second layer of parallelism.

@@

\section*{Results}

@@

\subsubsection*{Benchmarking Methodology}

@@

\subsection*{MaBoSS.GPU}

@@

\subsection*{MaBoSS.MPI}

@@

\section*{Conclusions}

In this work, we presented two new implementations of MaBoSS, a continuous time Boolean model simulator, for HPC environments: the first implementation, MaBoSS.GPU, is designed to exploit the computational power of massively parallel GPU hardware, and the second implementation, MaBoSS.MPI, which enables MaBoSS to scale to many nodes of HPC clusters. We evaluated the performance of these implementations on real-world and synthetic models and demonstrated that both implementations are capable of providing significant speedups over the original CPU implementation. The GPU implementation shows $145-326\times$ speedup on real-world models, and the MPI implementation delivers a close-to-linear strong scaling on big models.

@@

\subsection*{Future work}

@@

Finally, to enable the simulation of extra-large models with thousands of nodes, the two presented parallelization methods could be combined into a single high-performance solution by running GPU implementation on multiple MPI nodes.

\begin{acknowledgements}
We thank Gautier Stoll for his guidance and fruitful discussion.
\end{acknowledgements}

\begin{funding}
This work was supported by the European Commission under the PerMedCoE project [H2020-ICT-951773] and SVV project number 260 698.
\end{funding}

\section*{Bibliography}
\bibliography{bibliography}

%% You can use these special %TC: tags to ignore certain parts of the text.
%TC:ignore
%the command above ignores this section for word count
\onecolumn
\newpage

>>>>

\documentclass[a4paper,num-refs]{oup-contemporary}
% \RequirePackage[2020-02-02]{latexrelease}
% \documentclass[times, twoside]{zHenriquesLab-StyleBioRxiv}

\journal{gigascience}

\usepackage{graphicx}
\usepackage{siunitx}

%\usepackage{blindtext}
\usepackage{algorithm}
\usepackage{algcompatible}

\title{MaBoSS for HPC environments: Implementations of the continuous time Boolean model simulator for large CPU clusters and GPU accelerators}
% \shorttitle{MaBoSS for HPC environments}

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[1,\authfn{1}]{Adam Šmelko}
\author[2]{Miroslav Kratochvíl}
\author[3,4,5]{Emmanuel Barillot}
% \author[3,4,5]{Laurence Calzone}
\author[3,4,5,\authfn{1}]{Vincent Noël}

\affil[1]{Department of Distributed and Dependable Systems, Charles University, Prague, Czech Republic}
\affil[2]{Luxembourg Centre for Systems Biomedicine, University of Luxembourg, Esch-sur-Alzette, Luxembourg}
\affil[3]{Institut Curie, Université PSL, F-75005, Paris, France}
\affil[4]{INSERM, U900, F-75005, Paris, France}
\affil[5]{Mines ParisTech, Université PSL, F-75005, Paris, France}

\authnote{\authfn{1}smelko@d3s.mff.cuni.cz; vincent.noel@curie.fr}

\papercat{Paper}

\runningauthor{Šmelko et al.}

\jvolume{00}
\jnumber{0}
\jyear{2023}

@@

\begin{frontmatter}
\maketitle

@@

\begin{keywords}
Computational Biology | High Performance Computing | Boolean models
\end{keywords}
\end{frontmatter}

\section{Introduction}
Biological systems are large and complex, and understanding their behavior remains critical to designing new therapies for complex diseases such as cancer. A crucial approach in this endeavor is building computational models from existing knowledge and analyzing them to find intervention points and predict the efficacy of new treatments.
Many different frameworks have been used to describe biological systems, from quantitative systems of differential equations to more qualitative approaches such as Boolean models. While the former seems more adapted to represent complex behavior, such as non-linear dependencies, the latter is being increasingly used because it is capable of representing very large systems.
Many Boolean models have been built to describe biological systems to tackle a variety of problems: from understanding fundamental properties of cell cycle~\cite{faure2006cellcycle,sizek2019boolean} to more advanced properties of cancer~\cite{fumia_carcinogenesis_2013, montagud2022prostate}.

Historically, the task of building Boolean models involved reading an extensive amount of literature and summarizing it in a list of essential components and their interactions. More recently, thanks to advances in databases listing such interactions~\cite{licata2020signor,turei2016omnipath} and to experimental techniques providing information on a bigger number of components, the automatic methods have been designed to infer Boolean formulas from the constraints encoded in the knowledge and the experimental data~\cite{10.1093/bioinformatics/btaa484,chevalier2020synthesis,benevs2023boolean} to build large Boolean models. While this effort faces many challenges, we believe it is the best way to study the large-scale complexity of biological systems. However, in order to analyze the dynamic properties of such large Boolean models, we need to develop efficient simulation tools. In this effort, adapting existing software to modern HPC computing architectures is an important goal.

Our work focuses on extending MaBoSS~\cite{stoll2012continuous, stoll2017maboss} --- a stochastic Boolean simulator that performs estimations of state probability trajectories based on Markov chains. 
The main contributions of this paper are twofold:
\begin{itemize}
    \item MaBoSS.GPU, a GPU-accelerated implementation of MaBoSS, which is designed to exploit the computational power of massively parallel GPU hardware.
    \item MaBoSS.MPI, a parallel implementation of MaBoSS which can scale to multinode environments, such as large CPU clusters.
\end{itemize}

The source codes of the proposed implementations are publicly available at their respective GitHub repositories\footnote{\url{https://github.com/sysbio-curie/MaBoSS.GPU}} \footnote{\url{https://github.com/sysbio-curie/MaBoSS-env-2.0}}. We also provide the scripts, presented graphs, data and instructions to reproduce the benchmarks in the replication package\footnote{\url{https://github.com/asmelko/gigascience24-artifact}}.

The paper is organized as follows. In the next section, we discuss the MaBoSS algorithm and its complexity. In the following section, we detail the implementation of MaBoSS.GPU and MaBoSS.MPI. In the results section, we present the performance benchmarks of the new implementations. Finally, we conclude the paper and discuss the future work. [TODO - add section refs]

\section{Background}

@@

\section{Complexity evaluation and parallelization opportunities of MaBoSS}

@@

\subsection{Simulation complexity and parallelization}

@@

\subsection{Statistics computation complexity and parallelization}

@@

\section{Implementation}\label{sec:implementation}

@@

\subsection{MaBoSS.GPU}
Porting MaBoSS algorithm to GPUs poses an interesting task from the optimizations perspective. A simple rewrite to an annotation-based tool, such as OpenACC, may not be the most optimal strategy due to the usage of complex data structures, such as hash maps and expression trees, that are not well suited for GPU hardware. We divided the implementation into two parts: the bare \emph{simulation} and the trajectory \emph{statistics computation}.

\subsubsection{Simulation}

@@

\subsubsection{Statistics computation}

@@

\subsection{MaBoSS.MPI}
Since MaBoSS.MPI and MaBoSS.GPU projects were developed in parallel, the MPI implementation of MaBoSS is based on the original CPU code. Same as in GPU code, CPU implementation also assigns a single thread to a single trajectory. The MPI implementation simply extends this distribution of work among multiple MPI nodes, adding the second layer of parallelism.

@@

\section{Results}

@@

\subsection{Benchmarking Methodology}

@@

\subsection{MaBoSS.GPU}

@@

\subsection{MaBoSS.MPI}

@@

\section{Conclusions}

In this work, we presented two new implementations of MaBoSS tool, a continuous time Boolean model simulator, designed for HPC environments:  MaBoSS.GPU which is designed to exploit the computational power of massively parallel GPU hardware, and MaBoSS.MPI which enables MaBoSS to scale to many nodes of HPC clusters. We evaluated the performance of these implementations on real-world and synthetic models and demonstrated that both variants are capable of providing significant speedups over the original CPU code. The GPU implementation shows $145-326\times$ speedup on real-world models, and the MPI implementation delivers a close-to-linear strong scaling on big models.

@@

\subsection{Future work}

@@

Finally, to enable the simulation of extra-large models with thousands of nodes at billions of samples, the two presented parallelization methods could be combined into a single high-performance solution by running GPU implementation on multiple MPI nodes.

\section{Acknowledgements}
We thank Laurence Calzone and Gautier Stoll for their guidance and fruitful discussions.

\section{Funding}
This work was supported by the European Commission under the PerMedCoE project [H2020-ICT-951773] and SVV project number 260 698.

% \section*{Bibliography}
\bibliography{bibliography}

%% You can use these special %TC: tags to ignore certain parts of the text.
%TC:ignore
%the command above ignores this section for word count
% \onecolumn
% \newpage
